Algorithmic detection of conserved quantities of finite-difference schemes for partial differential equations

### 2023-05-02 Papers/Paper ACM/tmp/CLpaper.tex
Friedmann paper

\documentclass[manuscript]{acmart}

@@

\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[latin1]{inputenc}
\usepackage{hyperref}
\usepackage{dsfont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{ifthen}
\usepackage{mmacells}
\usepackage[invisible]{MLaTex}

%\usepackage[displaymath,mathlines]{lineno}
%\linenumbers
%\pagewiselinenumbers
%\modulolinenumbers[1]

%permite que formulas
%se dividam por duas
%paginas diferentes
\allowdisplaybreaks

%[hypertex]

%\usepackage{showkeys}
%\usepackage[norefs]{refcheck}

@@

%\usepackage{curves}

%\newcommand{\cal}{\mathcal}
\newcommand{\argmin}{\operatorname{argmin}}
\newcommand{\sech}{\operatorname{sech}}
\newcommand{\intr}{\operatorname{int}}
\newcommand{\arccosh}{\operatorname{arccosh}}
\newcommand{\arcsinh}{\operatorname{arcsinh}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\spn}{\operatorname{span}}
\newcommand{\arccotg}{\operatorname{arccotg}}
\newcommand{\inter}{\operatorname{int}}
\newcommand{\var}{\operatorname{Var}}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\car}{\operatorname{car}}
\newcommand{\cl}{\operatorname{cl}}
\newcommand{\co}{\operatorname{co}}
\newcommand{\arctanh}{\operatorname{arctanh}}
\newcommand{\esssup}{\operatorname{esssup}}
\newcommand{\essinf}{\operatorname{essinf}}
\renewcommand{\div}{\operatorname{div}}
\newcommand{\grad}{\operatorname{grad}}

@@

\newcommand{\epsi}{\varepsilon}
\def\d{{\rm d}}
\def\dx{{\rm d}x}
\def\dy{{\rm d}y}
\def\da{{\rm d}a}
\def\dt{{\rm d}t}
%\def\leq{\leqslant}
%\def\geq{\geqslant}

@@

%\def\contentsname{}
\def\refname{References}
\def\bibname{References}
\def\chaptername{\empty}
\def\figurename{Fig.}
\def\abstractname{Abstract}

@@

%\theoremstyle{definition}

@@

\author{Diogo A. Gomes}
\affiliation{%
	\institution{King Abdullah University of Science and Technology (KAUST)}
	\department[0]{AMCS/CEMSE Division}
	\city{Thuwal}
	\postcode{23955-6900}
	\country{Saudi Arabia}}
\email{diogo.gomes@kaust.edu.sa}
\author{Friedemann Krannich}
\affiliation{%
	\institution{King Abdullah University of Science and Technology (KAUST)}
	\department[0]{AMCS/CEMSE Division}
	\city{Thuwal}
	\postcode{23955-6900}
	\country{Saudi Arabia}}
\email{friedemann.krannich@kaust.edu.sa}
\author{Ricardo de Lima Ribeiro}
\affiliation{%
	\institution{King Abdullah University of Science and Technology (KAUST)}
	\department[0]{AMCS/CEMSE Division}
	\city{Thuwal}
	\postcode{23955-6900}
	\country{Saudi Arabia}}
\email{ricardo.ribeiro@kaust.edu.sa}

@@

\begin{abstract}
Many partial differential equations (PDEs) admit conserved quantities like mass or energy. Those quantities are often essential to establish well-posed results. When approximating a PDE by a finite-difference scheme, it is natural to ask whether related discretized quantities remain conserved over time by the scheme. Such conservation may establish the stability of the numerical scheme. We present an algorithm for checking the preservation of a polynomial quantity under a polynomial finite-difference scheme. Our schemes can be explicit or implicit, have higher-order time and space derivatives, and an arbitrary number of variables. 
Additionally, we present an algorithm for, given a scheme, finding conserved quantities. 
We illustrate our algorithm with several finite-difference schemes.
\end{abstract}

@@

\section{Introduction}
%\label{intro}
Many partial differential equations (PDEs) admit integral quantities, that are conserved in time. 
For example, the advection equation preserves energy and the heat equation conserves mass. 
When approximating PDEs by a finite-difference scheme, the question arises whether such quantities are conserved by the scheme. 
Such information can be crucial to estimate whether a scheme approximates a PDE accurately and to determine its stability.\\
Computations for determining conservation can get rather tedious.
Hence, automating them in computer algebra systems is desirable.\\
Conserved quantities correspond to conservation laws, admitted by the PDE. A conservation law \cite{cheviakov20} is an equation of the form
	D_t\Phi[u]+D_x\Psi[u]=0
holding for all solutions $u$ of the considered PDE,
that induces the conserved quantity $\int \Phi[u]\ dx$ as
	\frac{d}{dt}\int \Phi[u]\ dx=\int D_t\Phi[u]\ dx=-\int D_x\Psi[u]\ dx=0
assuming periodic boundary conditions in $x$.\\
An analog formulation describes conservation laws for schemes of PDEs, where the derivatives are replaced by shifts \cite{hydon01}.  The key for the construction of schemes, that admit certain conservation laws from the continuous PDE, is the use of the discrete Euler operator \cite{cheviakov20} (also called discrete variational derivative). Kupershmidt \cite[II. Theorem 31]{kupershmidt85} discovered, that an equation is a discrete conservation law if and only if it belongs to the kernel of the discrete Euler operator. Hence, standard approaches for constructing schemes with conservation are either discretizing the continuous conservation law or finding multipliers for the scheme \cite{dorodnitsyn21}, such that the result is in the kernel of the discrete Euler operator.
This strategy was, for example, recently used by Dorodnitsyn et al. to develop a scheme for the shallow water equation, that preserves energy \cite{dorodnitsyn21}. Cheviakov et al. used this idea to find schemes for the linear and nonlinear wave equation, that admit several discrete analogs of continous conservation laws \cite{cheviakov20}.
The same strategies for finding conservation laws, that are described above, work for continuous PDEs or semi-discretized PDEs:
Algorithms for finding conserved quantities, using multipliers and the Euler operator for PDEs were developed and implemented by Cheviakov \cite{cheviakov07,cheviakov10}.
Hereman et al. \cite{hereman04,hereman08}, proposed an algorithm to compute conserved densities for semi-discretized schemes for PDEs with first-order time derivative. 
Their algorithm uses the scaling symmetries of the scheme to construct conserved quantities and calculates their coefficients using the discrete Euler operator.
Gao et al. extended this algorithm to first-order time-explicit schemes \cite{gao02} and first-order time-implicit schemes \cite{gao04}.\\
Bihlo et al. \cite{bihlo12} used another approach and derived finite-difference schemes for the shallow water equations, that are symmetry-invariant on an adaptive mesh and then analyzed them regarding conserved quantities.\\
In this paper, we propose an algorithm that checks if a quantity is conserved in time under a scheme. Our approach differs from the ideas described above, as we do not try to construct discrete $\Phi$ and $\Psi$, using the scheme, but we check if a given discrete $\Phi$ is conserved in time under a given scheme.\\
Conserved quantities usually involve integral expressions or their discrete analog, sums.
Gomes et al. proposed algorithms for the simplification of sums \cite{gomes20}.
They also developed and implemented algorithms for detecting quantities conserved by PDEs and semi-discretized schemes. 
We revise these techniques in Section \ref{assp}. 
To generalize those methods to situations where we do not sum over all arguments of the functions involved, we introduce the discrete partial variational derivative (Section \ref{pvard}).
The main contribution of this paper is an algorithm for checking the conservation of a quantity under a numerical scheme.
Gerdt showed \cite{gerdt12}, that the quantity is conserved, if its discrete time derivative belongs to the difference ideal generated by the scheme. However, some quantities may add to a constant (e.g. telescopic sums) and thus be trivially preserved without belonging to the difference ideal. Moreover, Gerdt's algorithm may not terminate, as the Gr{\"o}bner basis for the difference ideal may not be finite.
We overcome these issues by combining the discrete partial variational derivative with a polynomial ideal (instead of a difference ideal) with finite Gr{\"o}bner basis (Section \ref{algodetec}). 
Our algorithm works for schemes that are explicit and implicit in time and can treat schemes with several higher-order time  and space derivatives, and with several space dimensions.
Additionally, we can handle systems of equations and schemes with parameters. 
We have implemented this algorithm as part of a package in {\sc Mathematica} \cite{wolfram21} \footnote{The code is available upon request. For access contact ricardo.ribeiro@kaust.edu.sa.}.
In the examples, we show that our code finds conserved quantities and proper schemes for the time-implicit and time-explicit discretization of the Burgers equation and a system of PDEs arising in the study of mean-field games (Section \ref{eaa}).

\section{Preliminaries}
\label{assp}
Here, we follow the ideas in \cite{gomes20}.
In this paper, subscripts denote indices related to coordinates or tuples and superscripts denote indices related to sequences.
To avoid boundary terms, we work with periodic functions in $\Zz^d$.
	Let $N$, $d$ and $m$ be positive integers. The \emph{discrete torus} is $\Ii:=\{0,\ldots,N-1\}^d\subset \Zz^d$. Define the space
		\mathcal{P}(\Ii,\Rr^m):=\{u\ |\ u:\Ii\rightarrow\Rr^m\},
	extended to $\Zz^d$ by periodicity.
%Assuming periodicity allows us to extend the functions in $\mathcal{P}(\Ii)$ to $\Zz^d$, and we avoid discussing boundary terms.
	The space of functionals $\mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$ (not necessarily linear) on $\mathcal{P}(\Ii,\Rr^m)$ is
	\mathcal{F}(\mathcal{P}(\Ii,\Rr^m)):= \Bigg\{\mathcal{J}:\mathcal{P}(\Ii,\Rr^m)\to \Rr\ \Bigg|\ \mathcal{J}[u]=\sum_{n\in\Ii}F_n[u],\\
	\text{ where }
	F_n[u]=G(u(n+e^1),u(n+e^2),\ldots,u(n+e^k))\\
	\text{ and }\{e^1,\ldots,e^k\}\subset\Zz^d,\ \text{G is smooth}
In the previous definition, the $e^1,\ldots,e^k$ are not necessarily unit vectors and can vary between functionals.

@@

	\label{functional1}
	Let $d=m=1$ and
	$	\mathcal{J}[u]:=\sum_{n}(u(n+1)-u(n))^2$.
To simplify sums, the discrete variational derivative is a useful tool.
Let $u,v\in\mathcal{P}(\Ii,\Rr^m)$, $\epsilon 
	\in\Rr$ and $\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$. Define
		D\mathcal{J}[u](v):=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}.
	Below in \eqref{periodicity}, we shift the indices due to periodicity of $u$ and $v$,
	\begin{gather}
	\notag	D\mathcal{J}[u](v)=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}=\frac{d}{d\epsilon}\sum_{n\in\Ii}F_n[u+\epsilon v]\bigg|_{\epsilon=0}\\
	\notag	=\sum_{n\in\Ii}\frac{d}{d\epsilon}G(u(n+e^1)+\epsilon v(n+e^1),\ldots,u(n+e^k)+\epsilon v(n+e^k))\bigg|_{\epsilon=0}\\
	\notag	=\sum_{n\in\Ii}D_1G(u(n+e^1),\ldots,u(n+e^k))v(n+e^1)\\
	\notag	+\ldots+D_kG(u(n+e^1),\ldots,u(n+e^k))v(n+e^k)\\
	\label{periodicity}	=\sum_{n\in\Ii}D_1G(u(n),u(n+e^2-e^1),\ldots,u(n+e^k-e^1))v(n)\\
	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))v(n)\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))v(n)\\
	\notag	=\sum_{n\in\Ii}\Big(D_1G(u(n),u(n+e^2-e^1),\ldots,u(n+e^k-e^1))\\
	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))\Big)v(n)\\
	\notag	=:\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u](n)v(n).	
	\end{gather}
$\mathcal{V}(\mathcal{J})$ is the \emph{discrete variational derivative} of $\mathcal{J}$.
Because we assume all related functions to be smooth, the discrete variational derivative always exists.
	Let us return to Example \ref{functional1} and compute
		D\mathcal{J}[u](v)=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}=\frac{d}{d\epsilon}\sum_{n\in\Ii}F_n[u+\epsilon v]\bigg|_{\epsilon=0}\\
		=\sum_{n\in\Ii}\frac{d}{d\epsilon}G[u(n+1)+\epsilon v(n+1),u(n)+\epsilon v(n)]\bigg|_{\epsilon=0}\\
		=\sum_{n\in\Ii}2(u(n+1)-u(n))v(n+1)+(-2)(u(n+1)-u(n))v(n)\\
		=\sum_{n\in\Ii}2(u(n)-u(n-1))v(n)+(-2)(u(n+1)-u(n))v(n)\\
		=\sum_{n\in\Ii}\Big(2(u(n)-u(n-1))+(-2)(u(n+1)-u(n))\Big)v(n)\\
		=\sum_{n\in\Ii}\Big(-2u(n+1)+4u(n)-2u(n-1)\Big)v(n)\\
		=\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u](n)v(n)
	and hence $\mathcal{V}(\mathcal{J})[u](n)=-2u(n+1)+4u(n)-2u(n-1)$.
The algorithms for the simplification of sums presented in \cite{gomes20} rely on the following result:
\begin{teo}\label{theo1}
	Let $\mathcal{J},\tilde{\mathcal{J}}\in \mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$. If $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[v]=0$ for all $v\in\mathcal{P}(\Ii,\Rr^m)$ and if there exists $u_0\in\mathcal{P}(\Ii,\Rr^m)$ such that $\mathcal{J}[u_0]=\tilde{\mathcal{J}}[u_0]$, then
		$\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$.
		 $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u]=0$
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$, if $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\begin{proof}
		(\mathcal{J}-\tilde{\mathcal{J}})[u]=(\mathcal{J}-\tilde{\mathcal{J}})[u]-(\mathcal{J}-\tilde{\mathcal{J}})[u_0]\\
		=\int_{0}^{1}\frac{d}{d\lambda}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)]\ d\lambda\\
		=\int_{0}^{1}\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)](u-u_0)\ d\lambda=0
	and hence $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{proof}
We use Theorem \ref{theo1} to examine if different sums represent the same quantity.
	Let $d=m=1$ and consider the functionals
		\mathcal{J}[u]:=\sum_{n\in\Ii}u(n)u(n+1)\text{ and }
		\tilde{\mathcal{J}}[u]:=\sum_{n\in\Ii}u(n-2)u(n-1).
	It is clear, that $\mathcal{J}$ and $\tilde{\mathcal{J}}$ represent the same quantity (by shifting $n$). We confirm this, using the discrete variational derivative:
		D(\mathcal{J}-\tilde{\mathcal{J}})[u](v)=\\
		=\sum_{n\in\Ii}(u(n+1)+u(n-1)-u(n+1)-u(n-1))v(n)=0.
	Hence, $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})=0$ and $\mathcal{J}[0]=0=\tilde{\mathcal{J}}[0]$. Therefore, both functionals sum to the same quantity.
	Let $d=2$ and $m=1$. It may not be obvious, that

@@

Let $n=(n_1,\ldots,n_d)$ be the variables for the functions in $\mathcal{P}(\Ii,\Rr^m)$. We call $n$ the space variables. Let $l\in\Nn$ and call $t:=(t_1,\ldots,t_l)\in\Nn^l$ the time variables. Later in this paper, we only consider the case $l=1$.

@@

		\Bigg\{\mathcal{J}:\mathcal{P}(\Ii\times\Nn^l)\rightarrow\mathcal{D}(\Nn^l) \Bigg|\ \mathcal{J}[u](t)=\sum_{n\in\Ii}F_{(n,t)}[u],\\

@@

Here, all quantities, whose conservation in time we check, are functions $\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l))$. %We restrict $G$ to the set of polynomials in $k$ variables, because our algorithm below can only treat polynomial quantities.
Let $\tilde{n}:=(n_1,\dots,n_d,s_1,\dots,s_l)\in\Ii\times\Nn^l$. Using the Kronecker delta $\delta$, we rewrite

@@

	%=\sum_{n\in\Ii}G(u(n+e^1),u(n+e^2),\ldots,u(n+e^k))\delta_{t_1,\ldots,t_{d-l+1}}(n_l,\ldots,n_d)

@@

		=:\sum_{\tilde{n}}\mathcal{V}^l(\mathcal{J})[u](\tilde{n},t)v(\tilde{n}).

@@

	We define $\mathcal{V}^l(\mathcal{J})$ as the \emph{discrete partial variational derivative} of $\mathcal{J}$.

@@

		=\sum_{n,s}(\delta_{t+1}(s)-\delta_t(s))v(n,s)=\sum_{n,s}\mathcal{V}^1(\mathcal{J})(n,s,t)v(n,s).

@@

	\begin{EXE}
		LABEL="dpvd1"
		(*CODE*)
		variables = Association["indVars"->{n,t},"depVars"->{u},"timeVars"->{t}]
		expression = u[n,t+1] - u[n,t]
		PartialDVarDOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}	

@@

		\input{executables/dpvd1ctt.ed.tex}

>>>>

\documentclass[sigconf,twocolumn]{acmart}

@@

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[latin1]{inputenc}
\usepackage{eurosym}
\usepackage{epsfig}
\usepackage{hyperref}
\usepackage{dsfont}

\usepackage[displaymath,mathlines]{lineno}

\allowdisplaybreaks

\usepackage{ifthen}

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}

\usepackage{mmacells}
\usepackage[invisible]{MLaTex}

@@

\newcommand{\argmin}{\operatorname{argmin}}
\newcommand{\sech}{\operatorname{sech}}
\newcommand{\intr}{\operatorname{int}}
\newcommand{\arccosh}{\operatorname{arccosh}}
\newcommand{\arcsinh}{\operatorname{arcsinh}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\spn}{\operatorname{span}}
\newcommand{\arccotg}{\operatorname{arccotg}}
\newcommand{\inter}{\operatorname{int}}
\newcommand{\var}{\operatorname{Var}}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\car}{\operatorname{car}}
\newcommand{\cl}{\operatorname{cl}}
\newcommand{\co}{\operatorname{co}}
\newcommand{\arctanh}{\operatorname{arctanh}}
\newcommand{\esssup}{\operatorname{esssup}}
\newcommand{\essinf}{\operatorname{essinf}}
\renewcommand{\div}{\operatorname{div}}
\newcommand{\grad}{\operatorname{grad}}

@@

\newcommand{\epsi}{\varepsilon}
\def\d{{\rm d}}
\def\dx{{\rm d}x}
\def\dy{{\rm d}y}
\def\da{{\rm d}a}
\def\dt{{\rm d}t}

@@

\def\refname{References}
\def\bibname{References}
\def\chaptername{\empty}
\def\figurename{Fig.}
\def\abstractname{Abstract}

@@

@@

\author{Diogo A. Gomes}
\affiliation{
	\institution{King Abdullah University of Science and Technology (KAUST)}
	\department[0]{CEMSE Division}
	\department[1]{AMCS Program}
	\city{Thuwal}
	\postcode{23955-6900}
	\country{Saudi Arabia}}
\email{diogo.gomes@kaust.edu.sa}
\author{Friedemann Krannich}
\affiliation{
	\institution{King Abdullah University of Science and Technology (KAUST)}
	\department[0]{CEMSE Division}
	\department[1]{AMCS Program}
	\city{Thuwal}
	\postcode{23955-6900}
	\country{Saudi Arabia}}
\email{friedemann.krannich@kaust.edu.sa}
\author{Ricardo de Lima Ribeiro}
\affiliation{
	\institution{King Abdullah University of Science and Technology (KAUST)}
	\department[0]{CEMSE Division}
	\department[1]{AMCS Program}
	\city{Thuwal}
	\postcode{23955-6900}
	\country{Saudi Arabia}}
\email{ricardo.ribeiro@kaust.edu.sa}

@@

\begin{abstract}
Many partial differential equations (PDEs) admit conserved quantities like mass or energy. Those quantities are often essential to establish well-posed results. When approximating a PDE by a finite-difference scheme, it is natural to ask whether related discretized quantities remain conserved under the scheme. Such conservation may establish the stability of the numerical scheme. We present an algorithm for checking the preservation of a polynomial quantity under a polynomial finite-difference scheme. In our algorithm, schemes can be explicit or implicit, have higher-order time and space derivatives, and an arbitrary number of variables. 
Additionally, we present an algorithm for, given a scheme, finding conserved quantities. 
We illustrate our algorithm by studying several finite-difference schemes.
\end{abstract}

@@

\section{Introduction}
Many partial differential equations (PDEs) admit integral quantities, that are conserved in time. 
For example, the advection equation preserves energy and the heat equation conserves mass. 
When approximating PDEs by a finite-difference scheme, the question arises whether such quantities are conserved by the scheme. 
Such information can be crucial to estimate whether a scheme approximates a PDE accurately and to determine its stability.\\
While computations for determining conservation are often easy in simple cases, they get rather tedious for more complicated schemes or quantities. 
Hence, automating such computations 
in computer algebra systems is desirable.\\
An algorithm for finding conserved quantities for (continuous) differential equations was developed and implemented  in {\sc Maple} by Cheviakov (see \cite{cheviakov07} and \cite{cheviakov10}).
Hereman et al. (see for example \cite{hereman04} and \cite{hereman08}), proposed an algorithm to compute conserved densities for semi-discretized schemes for PDEs with first-order time derivative. 
Their algorithm, which is implemented in a {\sc Mathematica} package, uses the scaling symmetries of the scheme to construct conserved quantities and calculates their coefficients by using the discrete variational derivative.
Gao et al. extended this algorithm to first-order time-explicit schemes \cite{gao02} and to first-order time-implicit schemes \cite{gao04}. 
They also implemented it in the computer algebra system {\sc Reduce}.\\
In this paper, we propose an algorithm that checks if a quantity is conserved in time under a scheme.
Conserved quantities usually involve integral expressions or their discrete analog, sums.
Gomes et al. proposed algorithms for the simplification of sums in \cite{gomes20}.
They also developed and implemented algorithms for detecting quantities conserved by PDEs and semi-discretized schemes. 
We revise these techniques in Section \ref{assp}. 
To generalize those methods to situations where we do not sum over all arguments of the functions involved, we introduce the discrete partial variational derivative (Section \ref{pvard}).
The main contribution of this paper is an algorithm for checking the conservation of a quantity under a numerical scheme.
Gerdt showed in \cite{gerdt12}, that if the discrete time derivative of the quantity belongs to the difference ideal generated by the scheme, the quantity is conserved. However, some quantities may add to a constant (e.g. telescopic sums) and thus be trivially preserved without belonging to the difference ideal. Moreover, Gerdt's algorithm may not terminate, as the Gr{\"o}bner basis for the difference ideal may not be finite.
We overcome this issues by combining the discrete partial variational derivative with a polynomial ideal (instead of a difference ideal) with finite Gr{\"o}bner basis (Section \ref{algodetec}). 
Our algorithm works for schemes that are explicit and implicit in time and can treat schemes with several and higher-order time  and space derivatives and with several space dimensions.
Additionally, we can handle systems of equations and schemes with parameters. 
We have implemented this algorithm as part of a package in {\sc Mathematica} \cite{wolfram21}.
In the examples, we show that our code finds conserved quantities and proper schemes for the time-implicit and time-explicit discretization of the Burgers equation and a system of PDEs arising in the study of mean-field games (Section \ref{eaa}).

\section{Preliminaries}
\label{assp}
Throughout this section, we follow the ideas and computations in \cite{gomes20}.
Here, subscripts denote indices related to coordinates or tuples while superscripts denote indices related to sequences and families.
To avoid dealing with boundary terms, we work with periodic functions in $\Zz^d$.
\begin{definition}
	Let $N$, $d$ and $m$ be natural numbers. The \emph{discrete torus} is $\Ii:=\{0,\ldots,N-1\}^d\subset \Zz^d$. Define the space
	\begin{gather*}
		\mathcal{P}(\Ii,\Rr^m):=\{u\ |\ u:\Ii\rightarrow\Rr^m\},
	\end{gather*}
	extended to $\Zz^d$ by periodicity.
\end{definition}
\begin{definition}
	The space of functionals $\mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$ (not necessarily linear) on $\mathcal{P}(\Ii,\Rr^m)$ is
\begin{gather*}
	\mathcal{F}(\mathcal{P}(\Ii,\Rr^m)):= \Bigg\{\mathcal{J}:\mathcal{P}(\Ii,\Rr^m)\to \Rr\ \Bigg|\ \mathcal{J}[u]=\sum_{n\in\Ii}F_n[u],\\
	\text{ where }
	F_n[u]=G(u(n+e^1),u(n+e^2),\ldots,u(n+e^k))\\
	\text{ and }\{e^1,\ldots,e^k\}\subset\Zz^d,\ \text{G is smooth}
		\Bigg\}.
\end{gather*}
\end{definition}
In the previous definition, the $e^1,\ldots,e^k$ are not necessarily unit vectors and can vary between functionals. Here, we allow for smooth $G$. However, later we require $G$ to be a polynomial function.
\begin{example}
	\label{functional1}
	Let $d=1$, $m=1$ and define
	\begin{gather*}
		\mathcal{J}[u]:=\sum_{n\in\Ii}(u(n+1)-u(n))^2.
	\end{gather*}
\end{example}
To simplify sums, the discrete variational derivative is a useful tool.
Let $u,v\in\mathcal{P}(\Ii,\Rr^m)$, $\epsilon 
	\in\Rr$ and $\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$. Define
	\begin{gather*}
		D\mathcal{J}[u](v):=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}.
	\end{gather*}
	In \eqref{periodicity}, we shift the indices due to periodicity of $u$ and $v$,
	\begin{gather}
	\notag	D\mathcal{J}[u](v)=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}=\frac{d}{d\epsilon}\sum_{n\in\Ii}F_n[u+\epsilon v]\bigg|_{\epsilon=0}\\
	\notag	=\sum_{n\in\Ii}\frac{d}{d\epsilon}G(u(n+e^1)+\epsilon v(n+e^1),\ldots,u(n+e^k)+\epsilon v(n+e^k))\bigg|_{\epsilon=0}\\
	\notag	=\sum_{n\in\Ii}D_1G(u(n+e^1),\ldots,u(n+e^k))v(n+e^1)\\
	\notag	+\ldots+D_kG(u(n+e^1),\ldots,u(n+e^k))v(n+e^k)\\
	\label{periodicity}	=\sum_{n\in\Ii}D_1G(u(n),u(n+e^2-e^1),\ldots,u(n+e^k-e^1))v(n)\\
	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))v(n)\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))v(n)\\
	\notag	=\sum_{n\in\Ii}\Big(D_1G(u(n),u(n+e^2-e^1)\ldots,u(n+e^k-e^1))\\
	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))\Big)v(n)\\
	\notag	=:\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u]v(n).	
	\end{gather}
\begin{definition}
$\mathcal{V}(\mathcal{J})$ is the discrete variational derivative of $\mathcal{J}$.
\end{definition}
Because we assume all related functions to be smooth, the discrete variational derivative always exists.
\begin{example}
	Let us return to Example \ref{functional1} and compute
	\begin{gather*}
		D\mathcal{J}[u](v)=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}=\frac{d}{d\epsilon}\sum_{n\in\Ii}F_n[u+\epsilon v]\bigg|_{\epsilon=0}\\
		=\sum_{n\in\Ii}\frac{d}{d\epsilon}G[u(n+1)+\epsilon v(n+1),u(n)+\epsilon v(n)]\bigg|_{\epsilon=0}\\
		=\sum_{n\in\Ii}2(u(n+1)-u(n))v(n+1)+(-2)(u(n+1)-u(n))v(n)\\
		=\sum_{n\in\Ii}2(u(n)-u(n-1))v(n)+(-2)(u(n+1)-u(n))v(n)\\
		=\sum_{n\in\Ii}\Big(2(u(n)-u(n-1))+(-2)(u(n+1)-u(n))\Big)v(n)\\
		=\sum_{n\in\Ii}\Big(-2u(n+1)+4u(n)-2u(n-1)\Big)v(n)\\
		=\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u]v(n)
	\end{gather*}
	and hence $\mathcal{V}(\mathcal{J})[u]=-2u(n+1)+4u(n)-2u(n-1)$.
\end{example}

\subsection*{Fundamental theorem}
The algorithms for the simplification of sums presented in \cite{gomes20} rely on the following result:
\begin{teo}\label{theo1}
	Let $\mathcal{J},\tilde{\mathcal{J}}\in \mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$. If $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[v]=0$ for all $v\in\mathcal{P}(\Ii,\Rr^m)$ and if there exists $u_0\in\mathcal{P}(\Ii,\Rr^m)$ such that $\mathcal{J}[u_0]=\tilde{\mathcal{J}}[u_0]$, then
	\begin{equation*}
		\mathcal{J}[u]=\tilde{\mathcal{J}}[u]
	\end{equation*}
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$.\\
	Conversely,
	\begin{equation*}
		 \mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u]=0
	\end{equation*}
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$, if $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{teo}
\begin{proof}
	\begin{gather*}
		(\mathcal{J}-\tilde{\mathcal{J}})[u]=(\mathcal{J}-\tilde{\mathcal{J}})[u]-(\mathcal{J}-\tilde{\mathcal{J}})[u_0]\\
		=\int_{0}^{1}\frac{d}{d\lambda}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)]\ d\lambda\\
		=\int_{0}^{1}\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)](u-u_0)\ d\lambda=0
	\end{gather*}
	and hence $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{proof}
We use Theorem \ref{theo1} to examine if different sums represent the same quantity.
\begin{example}
	Let $d=m=1$ and consider the functionals
	\begin{gather*}
		\mathcal{J}[u]:=\sum_{n\in\Ii}u(n)u(n+1),\\
		\tilde{\mathcal{J}}[u]:=\sum_{n\in\Ii}u(n-2)u(n-1).
	\end{gather*}
	It is clear, that $\mathcal{J}$ and $\tilde{\mathcal{J}}$ represent the same quantity (by shifting $n$). We confirm this, using the discrete variational derivative:
	\begin{gather*}
		D(\mathcal{J}-\tilde{\mathcal{J}})[u](v)=\\
		=\sum_{n\in\Ii}(u(n+1)+u(n-1)-u(n+1)-u(n-1))v(n)=0.
	\end{gather*}
	Hence, $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})=0$ and $\mathcal{J}[0]=0=\tilde{\mathcal{J}}[0]$. Therefore, both functionals sum to the same quantity.
\end{example}

@@

	Let $d=2$. It may not be obvious, that

@@

Let $n=(n_1,\ldots,n_d)$ be the variables for the functions in $\mathcal{P}(\Ii,\Rr^m)$. We call $n$ the space variables. Let $l\in\Nn$ and call $t=(t_1,\ldots,t_l)\in\Nn^l$ the time variables. Later in this paper, we only consider the case $l=1$.

@@

		\Bigg\{\mathcal{J}:\mathcal{P}(\Ii\times\Nn^l)\rightarrow\mathcal{D}(\Nn^l) \Bigg|\ \mathcal{J}[u]=\sum_{n}F_{(n,t)}[u],\\

@@

Throughout this paper, all quantities, whose conservation in time we check, are functions $\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l))$. 
Define $\tilde{n}:=(n_1,...,n_d,s_1,...,s_l)\in\Ii\times\Nn^l$. Using the Kronecker delta $\delta$, we rewrite

@@

@@

		=:\sum_{\tilde{n}}\mathcal{V}^l(\mathcal{J})[u](t_1,\ldots,t_{l})v(\tilde{n}).

@@

	We define $\mathcal{V}^l(\mathcal{J})$ as the discrete partial variational derivative of $\mathcal{J}$.

@@

		=\sum_{n,s}(\delta_{t+1}(s)-\delta_t(s))v(n,s)=\sum_{n,s}\mathcal{V}^1(\mathcal{J})(t)v(n,s).

@@

@@
